A Monte Carlo Synthetic-Acceleration Method for Solving the Thermal Radiation Diffusion Equation\tnoteref{ornl-cr}

### 2013-05-15 doc/jcomp2012/jcomp2012.tex
Added more results for marshak and dog-leg problem.

The following scheme will then converge in one iteration:
\begin{subequations}
  \begin{gather}
    \vx^{l+1} = (\vI - \vA)\vx^l + \vb\:,\\ \vA \delta \vx^{l+1} =
    \vr^{l+1}\;,
    \label{eq:synthetic-residual-solve}\\ 
    \vx=\vx^{l+1}+\delta\vx^{l+1}\:.
  \end{gather}
\end{subequations}
To accelerate the fixed-point method and avoid inverting $\vA$ directly in
Eq.~(\ref{eq:synthetic-residual-solve}), we can instead use the above ideas to
create a new iterative scheme that applies the Monte Carlo method to estimate
the solution error. With this in hand, the Fixed-Point Monte Carlo
Synthetic-Acceleration (MCSA) method can be defined as follows:
\begin{subequations}
  \label{eq:MCSA-iteration}
  \begin{gather}
    \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:,\\  \vr^{l+1/2} = \vb -
    \vA\vx^{l+1/2}\:,
    \label{eq:MCSA-residual}\\     
     \hat{\vA}\delta\vx^{l+1/2} = \vr^{l+1/2}\:,
    \label{eq:MCSA-residual_solve}\\ 
     \vx^{l+1}=\vx^{l+1/2}+\delta\vx^{l+1/2}\:.
  \end{gather}
\end{subequations}
The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that the
Monte Carlo solution only approximately inverts this operator.  Thus, we have
defined a scheme in which the initial estimate of $\vx$ in each iteration is
updated using a single fixed-point iteration\footnote{A subspace method could
  be used instead of a stationary method here.}.  The residual is calculated
and used as a source to estimate the error, $\delta\vx^{l+1/2}$, by solving
Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint method. The
error is then used to calculate an updated iterate of the solution vector,
$\vx^{l+1}$.  The entire sequence is iterated to convergence based on the
following stopping criterion~\cite{kelley_1995}:
\begin{equation}
  \|\vr\|_{\infty} < \epsilon\cdot\|\vb\|_\infty\:.
  \label{eq:stopping-criteria}
\end{equation}

@@

Here we examine the effectiveness of the MCSA method and compare it to
traditional deterministic solution methods.  The deterministic methods
considered are preconditioned Conjugate Gradient (PCG) and
preconditioned fixed-point (Richardson) iteration (PFIX).  The PCG
method uses right-left preconditioning with $\ve{M}=\ve{L}\ve{L}^{T}$,
where $\ve{L} = \ve{L}^{T} = \sqrt{\diag(\ve{D})}$.  Naturally, more
powerful preconditioning methods could be used; however, this provides
a fair comparison with the preconditioning used in the MCSA method.
Different preconditioning strategies within MCSA will be the focus of
a future paper.  The PFIX scheme is used to solve
Eq.~(\ref{eq:preconditioned-diffusion}), which is identical to the
system being solved with MCSA.  We examine each method on two 3D
problems: a Marshak wave problem and a multi-material duct problem.

@@

Figure~\ref{fig:marshak_10} shows the analytic Marshak Wave solution compared
to the MCSA solution.  As opposed to most Monte Carlo techniques, the MCSA
method generates solutions that are numerically equivalent to the PCG and PFIX
results.  Because the MCSA method converges the residual of the system, the
numerical accuracy is determined by the stopping criterion shown in
Eq.~(\ref{eq:stopping-criteria}).  Also, the MCSA method is not bound by the
Central Limit Theorem so convergence is not restricted to
$1/\sqrt{N}$~\cite{halton_1994,evans_2003}.
\begin{figure}[ht!]
  \centerline{ \includegraphics[width=5in,clip]{marshak_10.pdf}}
  \caption{ Solution of the Marshak wave problem at $t=100$~ns.  The analytic
    solution is compared to the MCSA solution.  The MCSA, PCG, and PFIXED
    methods are all converged to the same tolerance and therefore give
    numerically equivalent answers.}
  \label{fig:marshak_10}
\end{figure}

@@

Having analyzed the variation of the MCSA algorithm efficiency with number of
particles and weight cutoff, we now compare to traditional deterministic
algorithms.  Table~\ref{tab:marshak_comparison} shows timing results for the
Marshak problem using preconditioned CG (PCG), preconditioned Richardson
iteration (PFIX), and MCSA. The results show that the MCSA method is the most
efficient method, although the PCG results are nearly as fast.  More
importantly, the MCSA method clearly accelerates Richardson iteration, as
indicated by the fact that MCSA is 42\% faster than Richardson iteration
alone.
\begin{table}[ht!]
  \caption{Comparison of solution methods for the Marshak problem. All
    methods were run with a stopping criterion of $\sn{1}{-8}$.  The
    MCSA method used $N_p=10$ and $W_c = \sn{1}{-4}$.  The CPU time is
    measured relative to the MCSA method, which required 72.6 s on a
    3.6~GHz Xeon.}
  \label{tab:marshak_comparison}
  \begin{center}
    \begin{tabular}{lrr}\hline\hline
      \multicolumn{1}{c}{Method} & \multicolumn{1}{c}{Max Iterations}
      & \multicolumn{1}{c}{Relative CPU Time}\\\hline\hline  PCG &
      11 & 1.11 \\  PFIX & 11 & 1.42 \\  MCSA & 11 & 1.0
      \\ \hline\hline
    \end{tabular}
  \end{center}
\end{table}

@@

Table~\ref{tab:multimat_comparison} shows the timing results for the
multi-material problem using PCG, PFIX, and MCSA.  The results correspond
roughly with the results from the Marshak problem.  The MCSA is marginally
faster than PCG and 43\% faster than PFIX.
\begin{table}[htbp!]
  \caption{ Comparison of solution methods for the multi-material
    problem. All methods were run with a stopping criterion of
    $\sn{1}{-8}$.  The MCSA method used $N_p=1000$ and $W_c =
    \sn{1}{-3}$.  The problem was run to an elapsed time of 100~ns.
    The MCSA method required 3.4 hours on a 3.6~GHz Xeon.}
  \label{tab:multimat_comparison}
  \begin{center}
    \begin{tabular}{lrr}\hline\hline
      \multicolumn{1}{c}{Method} & \multicolumn{1}{c}{Max Iterations}
      & \multicolumn{1}{c}{Relative CPU Time}\\\hline\hline  PCG &
      18 & 1.03 \\  PFIX & 38 & 1.43 \\  MCSA & 20 & 1.0
      \\ \hline\hline
    \end{tabular}
  \end{center}
\end{table}

>>>>

The following scheme will then converge in one iteration:
\begin{subequations}
  \begin{gather}
    \vx^{l+1} = (\vI - \vA)\vx^l + \vb\:,\\ \vA \delta \vx^{l+1} =
    \vr^{l+1}\;,
    \label{eq:synthetic-residual-solve}\\
    \vx=\vx^{l+1}+\delta\vx^{l+1}\:.
  \end{gather}
\end{subequations}
To accelerate the fixed-point method and avoid inverting $\vA$ directly in
Eq.~(\ref{eq:synthetic-residual-solve}), we can instead use the above ideas to
create a new iterative scheme that applies the Monte Carlo method to estimate
the solution error. With this in hand, the Fixed-Point Monte Carlo
Synthetic-Acceleration (MCSA) method can be defined as follows:
\begin{subequations}
  \label{eq:MCSA-iteration}
  \begin{gather}
    \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:,\\  \vr^{l+1/2} = \vb -
    \vA\vx^{l+1/2}\:,
    \label{eq:MCSA-residual}\\
     \hat{\vA}\delta\vx^{l+1/2} = \vr^{l+1/2}\:,
    \label{eq:MCSA-residual_solve}\\
     \vx^{l+1}=\vx^{l+1/2}+\delta\vx^{l+1/2}\:.
  \end{gather}
\end{subequations}
The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that the
Monte Carlo solution only approximately inverts this operator.  Thus, we have
defined a scheme in which the initial estimate of $\vx$ in each iteration is
updated using a single fixed-point iteration\footnote{A subspace method could
  be used instead of a stationary method here.}.  The residual is calculated
and used as a source to estimate the error, $\delta\vx^{l+1/2}$, by solving
Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint method. The
error is then used to calculate an updated iterate of the solution vector,
$\vx^{l+1}$.  The entire sequence is iterated to convergence based on the
following stopping criterion~\cite{kelley_1995}:
\begin{equation}
  \|\vr\|_{2} < \epsilon\cdot\|\vb\|_2\:.
  \label{eq:stopping-criteria}
\end{equation}

@@

Here we examine the effectiveness of the MCSA method and compare it to
traditional deterministic solution methods.  We implemented Krylov subspace
methods to solve the radiation diffusion equation using the Trilinos solver
library \cite{1089021}.  Conjugate Gradient (CG) is that standard method for
symmetric positive-definite (SPD) systems like that defined in
Eq.~(\ref{eq:discrete-diffusion}).  However, for completeness we include some
comparisons using restarted GMRES($m$) so that asymmetric preconditioners can
be tested. The stopping criterion in Eq.~(\ref{eq:stopping-criteria}) is used
for all calculations. All methods are tested on two 3D problems: a
uniform-medium Marshak wave problem and a multi-material duct problem.

@@

Figure~\ref{fig:marshak_10} shows the analytic Marshak Wave solution compared
to the MCSA solution.  As opposed to most Monte Carlo techniques, the MCSA
method generates solutions that are numerically equivalent to the
deterministic results.  Because the MCSA method converges the residual of the
system, the numerical accuracy is determined by the stopping criterion shown
in Eq.~(\ref{eq:stopping-criteria}).  Also, the MCSA method is not bound by
the Central Limit Theorem so convergence is not restricted to
$1/\sqrt{N}$~\cite{halton_1994,evans_2003}.
\begin{figure}[ht!]
  \centerline{ \includegraphics[width=5in,clip]{marshak_10.pdf}}
  \caption{ Solution of the Marshak wave problem at $t=100$~ns.  The analytic
    solution is compared to the MCSA solution.  The MCSA, PCG, and PFIXED
    methods are all converged to the same tolerance and therefore give
    numerically equivalent answers.}
  \label{fig:marshak_10}
\end{figure}

@@

Having analyzed the variation of the MCSA algorithm efficiency with number of
particles and weight cutoff, we now compare to traditional deterministic
algorithms.  Table~\ref{tab:marshak_comparison} shows timing results for the
Marshak problem using CG with Jacobi and algebraic multigrid preconditioning
and GMRES with incomplete LU (ILU) and ILU with threshold cutoff (ILUT).  The
multigrid preconditioner is the smoothed aggregation algebraic multigrid
method implemented in the Trilinos ML package \cite{Gee:2007uj}.  In all the
problems that follow we use a $4/3$ damping factor in the aggregation and a
symmetric Gauss-Seidel smoother with 2 sweeps per level.

Analyzing Table~\ref{tab:marshak_comparison}, we see that the overall
iterations are more than a factor of 2 less for the Krylov solvers compared to
MCSA; however, the cost per iteration is signicantly more expensive.  For any
given problem, the cost per iteration is roughly linear with particle count as
illustrated in Fig.~\ref{fig:CPU_Np}.  Because this problem converges running
only a maximum of 10 particles per iteration, 156,464 total particles over
10,157 cycles resulting in an average of $\sim 1.6$ particles per iteration,
the cost per iteration is exceptionally small.  Running more particles per
iteration would reduce the number of iterations, but a greater cost per
iteration.
\begin{table}[ht!]
  \caption{Comparison of solution methods for the Marshak problem. All
    methods were run with a stopping criterion of $\sn{1}{-8}$.  The
    MCSA method used $N_p=10$ and $W_c = \sn{1}{-4}$.   Runtimes are relative
    to the MCSA performance that required 57.72 seconds on
    an 2.2 MHz AMD Opteron.}
  \label{tab:marshak_comparison}
  \begin{center}
    \small
    \begin{tabular}{llrrrr}\hline\hline
      \multicolumn{1}{c}{Method} &
      \multicolumn{1}{c}{Preconditioner} &
      \multicolumn{1}{c}{Total} &
      \multicolumn{1}{c}{Average} &
      \multicolumn{1}{c}{Relative Time} &
      \multicolumn{1}{c}{Relative}\\
      & & \multicolumn{1}{c}{Iterations} &
      \multicolumn{1}{c}{Iterations} &
      \multicolumn{1}{c}{per Iteration} &
      \multicolumn{1}{c}{Runtime} \\\hline\hline
      %%
      CG & Jacobi & 46,746 & 4.60 & 25.91 & 3.29 \\
      CG & ML & 12,150 & 1.20 & 74.72 & 2.69 \\
      GMRES & ILU & 40,294 & 3.97 & 103.02 & 9.19 \\
      GMRES & ILUT$^{\dagger}$ & 40,447 & 3.98 & 101.31 & 9.09 \\
      MCSA & Jacobi & 98,863 & 9.73 & 1.0 & 1.0
      %%
      \\ \hline\hline
      \multicolumn{6}{l}{
        $^{\dagger}$ The threshold cutoff was set to 0.01, and the fill was 1.}
    \end{tabular}
  \end{center}
\end{table}

@@

Table~\ref{tab:multimat_comparison} shows the timing results for the
multi-material problem using CG with Jacobi and ML preconditioning versus MCSA
run out to a time of 100~ns.  The MCSA method was set to use 1000 particles
per iteration resulting in 74,237,525 total particles over 10,157 cycles or
823.68 particles per iteration.  Because more particles are required to get a
reasonable estimate of the Neumann series, the relative efficiency per
iteration of MCSA versus CG is significanly reduced compared to the smaller
Marshak wave problem.  Nonetheless, MCSA is still marginally more efficient
than CG with Jacobi preconditioning and significantly more efficient than CG
with algegraic multigrid.
\begin{table}[htbp!]
  \caption{ Comparison of solution methods for the multi-material
    problem. All methods were run with a stopping criterion of
    $\sn{1}{-8}$.  The MCSA method used $N_p=1000$ and $W_c =
    \sn{1}{-3}$.  The problem was run to an elapsed time of 100~ns.
    Runtimes are relative to the MCSA performance that required 9,862.4
    seconds on a 2.2 MHz AMD Opteron.}
  \label{tab:multimat_comparison}
  \begin{center}
    \small
    \begin{tabular}{llrrrr}\hline\hline
      \multicolumn{1}{c}{Method} &
      \multicolumn{1}{c}{Preconditioner} &
      \multicolumn{1}{c}{Total} &
      \multicolumn{1}{c}{Average} &
      \multicolumn{1}{c}{Relative Time} &
      \multicolumn{1}{c}{Relative}\\
      & & \multicolumn{1}{c}{Iterations} &
      \multicolumn{1}{c}{Iterations} &
      \multicolumn{1}{c}{per Iteration} &
      \multicolumn{1}{c}{Runtime} \\\hline\hline
      %%
      CG & Jacobi & 42,881 & 4.22 & 2.93 & 1.02 \\
      CG & ML & 14,097 & 1.39 & 26.60 & 1.94 \\
      MCSA & Jacobi & 90,129 & 8.87 & 1.0 & 1.0
      %%
      \\ \hline\hline
    \end{tabular}
  \end{center}
\end{table}
